---
title: "Plotting functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(revolver)
```

Since the new release of `REVOLVER` (>= 0.3), we have implemented [all plotting functions](https://caravagn.github.io/revolver/reference/index.html)  using  [ggplot2](https://ggplot2.tidyverse.org/).

All functions now return either a `ggplot` object, or a more complext figure obtained by assembling `ggplot` objects either via `ggarrange` (package [ggpubr](https://cran.r-project.org/web/packages/ggpubr/index.html)) or `plot_grid` (package [cowplot](https://cran.r-project.org/web/packages/cowplot/)) `cowplot`. Notice that some functions might be only applicable after fits or clusters have been computed (the package will raise errors if you try to call a function on an object that does not have the required information to create the plot). 



In this vignette, we show the functions plotting  one of the cohort objects released in the `evoverse.datasets` [R package](https://github.com/caravagn/evoverse.datasets).
```{r, echo = TRUE, warning=FALSE}
# Data released in the 'evoverse.datasets'
data('TRACERx_NEJM_2017_REVOLVER', package = 'evoverse.datasets')
```

# Plotting cohort-level data

You can plot various information about the drivers in the cohort, and the trajectories detected in the cohort. See the manual of each one of these functions to understand non-trivial plots, and how to customize the input parameters.

```{r, echo = TRUE, warning=FALSE, fig.width=12, fig.height=9}
# Plot the occurrence of drivers across the cohort.
plot_drivers_occurrence(TRACERx_NEJM_2017_REVOLVER)
```


```{r, echo = TRUE, warning=FALSE, fig.width=6, fig.height=10}
# Plot the barplot of drivers' occurrences (clonal vs subclonal)
plot_drivers_clonality(TRACERx_NEJM_2017_REVOLVER) 
```

```{r, echo = TRUE, warning=FALSE, fig.width=13, fig.height=10}
# Plot the  graph-alike summary statistics for the cohort drivers.
plot_drivers_graph(TRACERx_NEJM_2017_REVOLVER)
```

This graph plot can be simplified reduce the entries used to generate the edges - try for instance using `min.occurrences = 5`.

You can also plot the penalty of the fit, and the DET index derived from that

```{r, echo = TRUE, warning=FALSE, fig.width=12, fig.height=10}
# Plot the index of Divergent Evolutionary Trajectories
plot_penalty(TRACERx_NEJM_2017_REVOLVER)
```


```{r, echo = TRUE, warning=FALSE, fig.width=12, fig.height=10}
# Plot the index of Divergent Evolutionary Trajectories
plot_DET_index(TRACERx_NEJM_2017_REVOLVER)
```


# Plotting patient's data

You can plot the data of a patient


# Plotting REVOLVER clusters and jackknife statistics

The results of using `REVOLVER` to cluster your cohort can be visualised with two functions

```{r, echo = TRUE, warning=FALSE, fig.width=10, fig.height=10}
# Plot the heatmaps of REVOLVER"s clusters, as tiles.
plot_clusters(TRACERx_NEJM_2017_REVOLVER)
```

The top tile is patients vs trajectories, and bottom is patients vs drivers. For drivers colours reflect mean CCF/ binary values of a driver in every patient, and clonality status. For trajectories colours reflect if they are initiating or progressing, depending on the present of GL in the trajectory (GL stands for germline). 

The dendrogram of the clusters can be visualised - notice the colours and patients' ordering that are matched in these two functions.
```{r, echo = TRUE, warning=FALSE, fig.width=10, fig.height=4}
# Plot the dendrogram of REVOLVER"s clusters
plot_dendrogram(TRACERx_NEJM_2017_REVOLVER)
```

After jackknife, you can also plot the co-clustering probability of the patients, the clustering stability and other statistics about the trajectories.

```{r, echo = TRUE, warning=FALSE, fig.width=12, fig.height=12}
# Plot the patients' jackknife co-clustering probability
plot_jackknife_coclustering(TRACERx_NEJM_2017_REVOLVER)
```

```{r, echo = TRUE, warning=FALSE, fig.width=8, fig.height=4}
# Plot the patients' jackknife cluster staability
plot_jackknife_cluster_stability(TRACERx_NEJM_2017_REVOLVER)
```

```{r, echo = TRUE, warning=FALSE, fig.width=8, fig.height=4}
# Plot the patients' jackknife trajectory frequencies.
plot_jackknife_trajectories_stability(TRACERx_NEJM_2017_REVOLVER)
```

 


# S3 plotting functions
